% A LaTeX template for MSc Thesis submissions to 
% Politecnico di Milano (PoliMi) - School of Industrial and Information Engineering
%
% S. Bonetti, A. Gruttadauria, G. Mescolini, A. Zingaro
% e-mail: template-tesi-ingind@polimi.it
%
% Last Revision: October 2021
%
% Copyright 2021 Politecnico di Milano, Italy. NC-BY

\documentclass{Configuration_Files/PoliMi3i_thesis}

%------------------------------------------------------------------------------
%	REQUIRED PACKAGES AND  CONFIGURATIONS
%------------------------------------------------------------------------------

% CONFIGURATIONS
\usepackage{parskip} % For paragraph layout
\usepackage{setspace} % For using single or double spacing
\usepackage{emptypage} % To insert empty pages
\usepackage{multicol} % To write in multiple columns (executive summary)
\setlength\columnsep{15pt} % Column separation in executive summary
\setlength\parindent{0pt} % Indentation
\raggedbottom  

% PACKAGES FOR TITLES
\usepackage{titlesec}
% \titlespacing{\section}{left spacing}{before spacing}{after spacing}
\titlespacing{\section}{0pt}{3.3ex}{2ex}
\titlespacing{\subsection}{0pt}{3.3ex}{1.65ex}
\titlespacing{\subsubsection}{0pt}{3.3ex}{1ex}
\usepackage{color}

% PACKAGES FOR LANGUAGE AND FONT
\usepackage[english]{babel} % The document is in English  
\usepackage[utf8]{inputenc} % UTF8 encoding
\usepackage[T1]{fontenc} % Font encoding
\usepackage[11pt]{moresize} % Big fonts

% PACKAGES FOR IMAGES
\usepackage{graphicx}
\usepackage{transparent} % Enables transparent images
\usepackage{eso-pic} % For the background picture on the title page
\usepackage{subfig} % Numbered and caption subfigures using \subfloat.
\usepackage{tikz} % A package for high-quality hand-made figures.
\usetikzlibrary{}
\graphicspath{{./Images/}} % Directory of the images
\usepackage{caption} % Coloured captions
\usepackage{xcolor} % Coloured captions
\usepackage{amsthm,thmtools,xcolor} % Coloured "Theorem"
\usepackage{float}

% STANDARD MATH PACKAGES
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{bm}
\usepackage[overload]{empheq} % For braced-style systems of equations.
\usepackage{fix-cm} % To override original LaTeX restrictions on sizes

% PACKAGES FOR TABLES
\usepackage{tabularx}
\usepackage{longtable} % Tables that can span several pages
\usepackage{colortbl}

% PACKAGES FOR ALGORITHMS (PSEUDO-CODE)
\usepackage{algorithm}
\usepackage{algorithmic}

% PACKAGES FOR REFERENCES & BIBLIOGRAPHY
\usepackage[colorlinks=true,linkcolor=black,anchorcolor=black,citecolor=black,filecolor=black,menucolor=black,runcolor=black,urlcolor=black]{hyperref} % Adds clickable links at references
\usepackage{cleveref}
\usepackage[square, numbers, sort&compress]{natbib} % Square brackets, citing references with numbers, citations sorted by appearance in the text and compressed
\bibliographystyle{unsrtnat} % You may use a different style adapted to your field

% OTHER PACKAGES
\usepackage{pdfpages} % To include a pdf file
\usepackage{afterpage}
\usepackage{lipsum} % DUMMY PACKAGE
\usepackage{fancyhdr} % For the headers
\fancyhf{}

% Input of configuration file. Do not change config.tex file unless you really know what you are doing. 
\input{Configuration_Files/config}

%----------------------------------------------------------------------------
%	NEW COMMANDS DEFINED
%----------------------------------------------------------------------------

% EXAMPLES OF NEW COMMANDS
\newcommand{\bea}{\begin{eqnarray}} % Shortcut for equation arrays
\newcommand{\eea}{\end{eqnarray}}
\newcommand{\e}[1]{\times 10^{#1}}  % Powers of 10 notation

%----------------------------------------------------------------------------
%	ADD YOUR PACKAGES (be careful of package interaction)
%----------------------------------------------------------------------------

%----------------------------------------------------------------------------
%	ADD YOUR DEFINITIONS AND COMMANDS (be careful of existing commands)
%----------------------------------------------------------------------------

%----------------------------------------------------------------------------
%	BEGIN OF YOUR DOCUMENT
%----------------------------------------------------------------------------

\begin{document}

\fancypagestyle{plain}{%
\fancyhf{} % Clear all header and footer fields
\fancyhead[RO,RE]{\thepage} %RO=right odd, RE=right even
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}}

%----------------------------------------------------------------------------
%	TITLE PAGE
%----------------------------------------------------------------------------

\pagestyle{empty} % No page numbers
\frontmatter % Use roman page numbering style (i, ii, iii, iv...) for the preamble pages

\puttitle{
	title=Investigating Changes in White Matter Microstructure in Chemobrain Using Diffusion MRI Representations, % Title of the thesis
	name=Petra Cipolla, % Author Name and Surname
	course= Biomedical Engineering - Ingegneria Biomedica, % Study Programme (in Italian)
	ID  = 10698521,  % Student ID number (numero di matricola)
	advisor= Prof. Giuseppe Baselli, % Supervisor name
	coadvisor={Prof. Daan Chrisitaens, Siebe Leysen}, % Co-Supervisor name, remove this line if there is none
	academicyear={2024-25},  % Academic Year
} % These info will be put into your Title page 

%----------------------------------------------------------------------------
%	PREAMBLE PAGES: ABSTRACT (inglese e italiano), EXECUTIVE SUMMARY
%----------------------------------------------------------------------------
\startpreamble
\setcounter{page}{1} % Set page counter to 1

% ABSTRACT IN ENGLISH
\chapter*{Abstract} 

Diffusion MRI offers a non-invasive method to study white matter (WM) microstructure by probing the motion of water molecules. Constrained Spherical Deconvolution (CSD) effectively extracts WM fibre orientation information from diffusion data by disentangling the contributions of different fibre bundles within a voxel by estimating the Orientation Distribution Function (ODF). Compared to the state-of-the-art method Multi-Tissue CSD (MT-CSD), which uses pre-defined tissue-specific response functions, Local Response Function Estimation in Spherical Deconvolution (LoRE-SD) estimates response functions locally within each voxel, resulting in unit-normalized ODFs. This normalization allows different scalings to be applied, each potentially emphasizing distinct tissue compartments. 
From the ODFs, we can extract fixels, each representing a population of fibres within a voxel running in the same direction. Fixel-Based Analysis (FBA) is a framework that assigns metrics to these fixels and enables group comparisons.
In this work, we used FBA to study chemobrain, a condition affecting chemotherapy-treated cancer survivors, who often experience cognitive impairment. Our goal was to investigate whether LoRE-SD could be used effectively in a FBA and identify potential differences between chemotherapy-treated breast cancer patients and control groups. ODFs were found using both MT-CSD and LoRE-SD, with multiple scaling strategies applied to the latter to emphasize intra-axonal content, anisotropy, or WM volume fraction. The results showed that the significant findings varied depending on the ODF modulation used with LoRE-SD, highlighting its flexibility in revealing different microstructural features. Notably, the fornix showed a significant increase in apparent fibre density in patients compared to controls, possibly reflecting compensatory mechanisms or network reorganization in response to chemotherapy-induced neuroinflammation. However, its thinness and proximity to the ventricles raise the possibility of CSF partial volume effects or registration errors. This thesis highlights the potential of LoRE-SD and FBA to study WM alterations, though interpretation remains difficult without a clearer understanding of the pathology.
\\
\\
\textbf{Keywords:} Diffusion MRI, White matter microstructure, Constrained Spherical Deconvolution, Chemobrain % Keywords

% ABSTRACT IN ITALIAN
\chapter*{Abstract in lingua italiana}
La risonanza magnetica pesata in diffusione offre un metodo non invasivo per studiare la microstruttura della sostanza bianca analizzando il movimento delle molecole d’acqua. Il metodo Constrained Spherical Deconvolution (CSD) separa il contributo di diversi fasci all’interno di un voxel, stimando l'Orientation Distribution Function (ODF), che descrive l’orientamento spaziale delle fibre nervose. Rispetto allo stato dell'arte Multi-Tissue CSD (MT-CSD), che utilizza funzioni di risposta predefinite per i tessuti, LoRE-SD stima la funzione di risposta localmente in ogni voxel, producendo ODF normalizzate scalabili per enfatizzare diversi compartimenti tissutali.
Dalle ODF si estraggono i fixel, ciascuno rappresentante una popolazione di fibre orientate nella stessa direzione. La Fixel-Based Analysis (FBA) assegna metriche ai fixel e consente confronti tra gruppi. 
In questa tesi, FBA è stata applicata allo studio del chemobrain, ovvero i sintomi cognitivi che possono manifestarsi dopo la chemioterapia. L’obiettivo era valutare per la prima volta LoRE-SD in un’analisi FBA e indagare differenze tra pazienti con tumore al seno trattate con chemioterapia e gruppi di controllo. Le ODF sono state calcolate con MT-CSD e LoRE-SD, utilizzando per quest’ultimo strategie di scalatura per enfatizzare contenuto intra-assonale, anisotropia o frazione di sostanza bianca. I risultati hanno mostrato che le regioni significative dipendono dal metodo di derivazione delle ODF e dal tipo di modulazione applicata, evidenziando la flessibilità di LoRE-SD.
In particolare, la fornice è risultata una regione con un aumento significativo della metrica Apparent Fibre Density nelle pazienti rispetto ai controlli. Questo effetto potrebbe riflettere meccanismi compensatori o una riorganizzazione dei circuiti nervosi in risposta alla neuroinfiammazione indotta dalla chemioterapia. Tuttavia, la fornice è una struttura molto sottile e vicina ai ventricoli, quindi soggetta a effetti di volume parziale con il liquido cerebrospinale e ad errori di registrazione.
La tesi dimostra il potenziale di LoRE-SD e FBA nello studio della sostanza bianca, pur evidenziando che l’interpretazione dei risultati resta limitata dalla scarsa conoscenza dei meccanismi patologici coinvolti.
\\
\\
\textbf{Parole chiave:} Risonanza magnetica pesata in diffusione, Microstruttura della sostanza bianca, Constrained Spherical Deconvolution, Chemobrain % Keywords (italian)

%----------------------------------------------------------------------------
%	LIST OF CONTENTS/FIGURES/TABLES/SYMBOLS
%----------------------------------------------------------------------------

% TABLE OF CONTENTS
\thispagestyle{empty}
\tableofcontents % Table of contents 
\thispagestyle{empty}
\cleardoublepage

%-------------------------------------------------------------------------
%	THESIS MAIN TEXT
%-------------------------------------------------------------------------
% In the main text of your thesis you can write the chapters in two different ways:
%
%(1) As presented in this template you can write:
%    \chapter{Title of the chapter}
%    *body of the chapter*
%
%(2) You can write your chapter in a separated .tex file and then include it in the main file with the following command:
%    \chapter{Title of the chapter}
%    \input{chapter_file.tex}
%
% Especially for long thesis, we recommend you the second option.

\addtocontents{toc}{\vspace{2em}} % Add a gap in the Contents, for aesthetics
\mainmatter % Begin numeric (1,2,3...) page numbering

% --------------------------------------------------------------------------
% NUMBERED CHAPTERS % Regular chapters following
% --------------------------------------------------------------------------
\chapter*{Introduction}
\input{introduction.tex}


\chapter{Overview}
\input{chapter_1.tex}



\chapter{Methods}
\input{chapter_2.tex}

\chapter{Results}
\input{chapter_3.tex}

\chapter{Discussion}
\input{chapter_4.tex}

\chapter{Conclusion}
\label{ch:conclusions}%
\input{conclusion.tex}

%-------------------------------------------------------------------------
%	BIBLIOGRAPHY
%-------------------------------------------------------------------------

\addtocontents{toc}{\vspace{2em}} % Add a gap in the Contents, for aesthetics
\bibliography{Thesis_bibliography} % The references information are stored in the file named "Thesis_bibliography.bib"

%-------------------------------------------------------------------------
%	APPENDICES
%-------------------------------------------------------------------------

\cleardoublepage
\addtocontents{toc}{\vspace{2em}} % Add a gap in the Contents, for aesthetics



% LIST OF FIGURES
\listoffigures

% LIST OF TABLES
\listoftables

% LIST OF SYMBOLS
% Write out the List of Symbols in this page
\chapter*{List of Abbreviations} % You have to include a chapter for your list of symbols (
\begin{table}[H]
    \centering
    \begin{tabular}{ll}
       % \textbf{Variable} & \textbf{Description} \\\hline\\[-9px]
        %$\bm{u}$ & solid displacement & m \\[2px]
        %$\bm{u}_f$ & fluid displacement & m \\[2px]
        ADC & Apparent Diffusion Coefficient\\
    AFD & Apparent Fibre Density\\
    BBB   & Blood-Brain Barrier \\
    CC & Corpus Callosum\\
    CFE & Connectivity-based Fixel Enhancement\\
    CM & Chemotherapy-Na\"ive Patients\\
    CNN & Convolutional Neural Network\\
    CP & Chemotherapy Patients\\
    CSD & Constrained Spherical Deconvolution\\
    CSF & Cerebrospinal Fluid\\
    dMRI   & Diffusion-weighted Magnetic Resonance Imaging \\
    DTI & Diffusion Tensor Imaging\\
    GM & Gray Matter \\
    FA & Fractional Anisotropy\\
    FBA & Fixel-Based Analysis\\
    FC & Fibre Cross-section\\
    FDC & Fibre Density and Cross-section\\
    FWE & Family-wise Error\\
    GLM & General Linear Model\\
    HARDI & High Angular Resolution Diffusion Imaging\\
    HC & Healthy Controls\\
    LoRE-SD & Local Response function Estimation in Spherical Deconvolution\\
    MD & Mean Diffusivity\\
   MT-CSD & Multi-tissue Constrained Spherical Deconvolution\\
   ODF & Orientation Distribution Function\\
   ROI & Region of Interest\\
   SH & Spherical Harmonics\\
   SNR & Signal-to-Noise Ratio\\
   TDI & Track-Density Imaging\\
      WM  & White Matter \\
      VBA & Voxel-Based Analysis
    \end{tabular}
\end{table}

% ACKNOWLEDGEMENTS
\chapter*{Acknowledgements}


\cleardoublepage

\end{document}
